\chapter{Conclusion and Outlook}

During the project described in this thesis and the attached papers, I have implemented a method to determine the structure of several small proteins using their experimental chemical shifts.
The structure of the CI-2 protein was solved in using only computer resources that are available in any lab, with only chemical shift data that was automatically recorded and assigned.

Lastly, I have attempted to fold several protein structures around 200 amino acids.
Out of 8 proteins greater than 100 residues, a good structure was located in four cases, out of which two were larger than 200 residues.
The last four likely failed due to inefficient use of the NOE restraints.
Since the existing code to handle NOE restraints in PHAISTOS did not perform well on large structures, I implemented a new NOE energy term, and this was used to fold the Rhodopsin structure (225 amino acids) to a CA-RMSD of 2.5 \AA~from the experimental X-ray structure using a set only 195 NOE data restraints and assigned backbone chemical shifts.
The same code was able to fold the Savinase structure to a CA-RMSD of 2.9 \AA~from the experimental X-ray structure using only distance restraints derived from evolutionary data and assigned chemical shifts.

This required implementing a version of CamShift, from scratch, in PHAISTOS, and implemented useful energy function rigorously founded in Bayesian statistics.
To aid the setup of calculations, a graphical user interface for PHAISTOS was created.
I have parametrized and implemented a version of ProCS to calculate amide proton chemical shifts, and shown that this parametrization yields structure that are in better agreement with experimental data than simulations using a chemical shift predictor parametrized from experimental data.
Furthermore, I have parametrized parts of the backbone atom ProCS chemical shift predictor and implemented this in a PHAISTOS module.
This required the implementation of FragBuilder Python API which was used to automatically setup, run, and collect data from more than 2,000,000 QM calculations.

The speed of the cached version of the backbone atom ProCS chemical shift predictor will allow an energy function based on chemical shifts to be included in simulations on proteins > 200 residues. Based on results obtained on the ENHD, Protein G and CI-2 proteins, this which will dramatically increase the accuracy of the energy function.

A newly developed NOE energy function shows encouraging results on folding of large structures, and further development of this module is promising.


In conclusion, I have demonstrated, that chemical shifts and sparse NOE data can, in some cases, be used with higher computational efficiency in PHAISTOS, than any other competing method. I have determined a protein structure in less than two days using automatically collected chemical shift data with computational resource available to any lab.
Lastly, I have folded some of the larges protein structures ever folded using similar approaches, while using only modest amount computational resources.
