\chapter*{Publication list}
\addcontentsline{toc}{chapter}{Publication list}

\section*{List of publications:}
\begin{enumerate}
    \item \underline{Anders S. Christensen}, Stephan P. A. Sauer, Jan H. Jensen (2011) Definitive benchmark study of ring current effects on amide proton chemical shifts. \textit{Journal of Chemical Theory and Computation}, 7:2078-2084.
    \item Wouter Boomsma, Jes Frellsen, Tim Harder, Sandro Bottaro, Kristoffer E. Johansson, Pengfei Tian, Kasper Stovgaard, Christian Andreetta, Simon Olsson, Jan B. Valentin, Lubomir D. Antonov, \underline{Anders S. Christensen}, Mikael Borg, Jan H. Jensen, Kresten Lindorff-Larsen, Jesper Ferkinghoff-Borg, Thomas Hamelryck (2013) PHAISTOS: A framework for Markov chain Monte Carlo simulation and inference of protein structure. \textit{Journal of Computational Chemistry}, 34:1697-1705.
    \item  \underline{Anders S. Christensen}, Troels E. Linnet, Mikael Borg, Wouter Boomsma, Kresten Lindorff-Larsen, Thomas Hamelryck, Jan H. Jensen (2013)  Protein Structure Validation and Refinement Using Amide Proton Chemical Shifts Derived from Quantum Mechanics. \textit{PLoS ONE} 8:e84123.
    \item \underline{Anders S. Christensen}, Thomas Hamelryck, Jan H. Jensen (2014) FragBuilder: An efficient Python library to setup quantum chemistry calculations on peptides models. \textit{PeerJ} 2:e277.
%    \item \underline{Anders S. Christensen}, Lars Bratholm, Simon Olsson, Thomas Hamelryck, Jan H. Jensen (2014) Weighting of chemical shift evidence in Monte Carlo simulation of proteins. \textit{ShareLatex} (unpublished).
%    \item Torus-DBN-CS-LARS
%    \item J-coupling Casper/Kongen
\end{enumerate}
\section*{List of public code:}
\begin{enumerate}
    \item FragBuilder (BSD license) \url{https://github.com/jensengroup/fragbuilder}
    \item CamShift module (BSD license) \url{https://github.com/jensengroup/camshift-phaistos}
    \item ProCS module (BSD license) \url{https://github.com/jensengroup/procs-phaistos}
    \item PHAISTOS (GPL license) \url{http://sourceforge.net/projects/phaistos/}
    \item GAMESS patch FMO-RHF:MP2 (GAMESS license/free) \url{https://github.com/andersx/fmo-rhf-mp2}
    \item PHAISTOS GUI (BSD license) \url{https://github.com/andersx/guistos}
    \item NOE module (BSD license) \url{https://github.com/andersx/noe-way-jose}
\end{enumerate}
\clearpage
\section*{List of other publications:}
\begin{enumerate}
    \item Casper Steinmann, Kristoffer L. Blædel, \underline{Anders S. Christensen}, Jan H. Jensen (2013) Interface of the polarizable continuum model of solvation with semi-empirical methods in the GAMESS program. \textit{PLoS ONE} 8:e67725.
    \item \underline{Anders S. Christensen}, Casper Steinmann, Dmitri G. Fedorov, Jan H. Jensen (2013) Hybrid RHF/MP2 geometry optimizations with the Effective Fragment Molecular Orbital Method. \textit{PLoS ONE} 9:e88800
    \item Jimmy C. Kromann, \underline{Anders S. Christensen}, Casper Steinmann, Martin Korth, Jan H. Jensen (2014) A third-generation dispersion and third-generation hydrogen bonding corrected PM6 method: PM6-D3H+. \textit{PeerJ PrePrints} 2:e353v1.
%    \item HF-3c Jimmy
%    \item PM6 Jimmy
%    \item h-bond Jimmy
\end{enumerate}

